Emulation of high multiplicity matrix elements for electron-positron and hadron-hadron colliders

### 2022-11-17 paper1.tex
update chapter 5

\section{Introduction}
\label{sec:intro}
Neural networks (NN) first appeared in the field of particle physics as a tool for event discrimination in the analysis of collider data. 
Since then neural networks and other machine learning techniques have proved useful in many other areas of the field. 
On the theory prediction side they have been used 
to improve the efficiency of Monte Carlo sampling \cite{Bendavid:2017zhk,Klimek:2018mza,Bothmann:2020ywa,Verheyen:2020bjw,Chen_2021}, 
to accelerate the simulation of radiation within a jet~\cite{Carrazza:2019cnt,Bothmann:2018trh,Dohi:2020eda}, 
to streamline the processes of generation and unweighting of simulated event samples 
\cite{Gao:2020zvv,Otten:2018kum,Hashemi:2019fkn,DiSipio:2019imz,Butter:2019cae,Bishara:2019iwh,Backes:2020vka,Butter:2020qhk,Alanazi:2020klf,Nachman:2020fff}.
Closer to the experimental measurements 
they have also been used to emulate detector simulation \cite{Paganini:2017dwg,SHiP:2019gcl,Derkach:2019qfk,Alanazi:2020jod},
 they can be used to perform unfolding \cite{Andreassen:2019cjw} or correcting for detector effects \cite{Bellagente:2019uyp},
and perform pileup subtraction \cite{Andreassen:2019cjw}.

The capacity of neural networks to approximate intricate functions have already been 
used to provide fast calculations of production cross-sections~\cite{Otten:2018kum,Buckley:2020bxg}.
In this article we investigate whether NNs can approximate production cross-sections more differentially by replacing computationally expensive 
matrix element calculations. 
The challenge with this endeavour is that matrix elements are plagued with numerous divergences that arise from infrared divergences.
In previous works \cite{Badger:2020uow,Aylett-Bullock:2021hmo} a combination of individual neural networks were used 
to approximate matrix elements. In order to deal with the complex structure of the matrix elements 
the authors of these studies divided the phase-space into sectors according to the infrared singularities and trained networks on these sectors,
thereby limiting the complexity of the fit by isolating a single divergence per sector. All sectors were then combined to make the final prediction.
While the authors of the initial study for electron-positron annihilation showed good agreement between the total cross-section and histogrammed distributions both at LO and NLO,
we note that the accuracy of the interpolation at the level of individual points was a lot worse than when averaged in the 
histograms. In addition, the performance of the extrapolation outside of its training phase-space 
(i.e. more singular configurations than those considered to fit the model) is problematic.

In this article we present a different approach to the emulation of matrix elements that incorporates the factorisation 
properties of the matrix elements in the interpolation model and therefore is able to safely 
extrapolate the matrix element in regions more singular than that covered by the data used for the training of the emulator. We find
that our interpolator also displays a much improved pointwise accuracy.

The article is organised as follows. 
Section~\ref{sec:fitting} introduces our factorisation-aware 
deep neural network model, Section~\ref{sec:results} showcases our results with a comparison with the 
model from Ref.~\cite{Badger:2020uow} in Section~\ref{sec:n3jet_comparison}, and an analysis of the performance of our model in Section~\ref{sec:mainresult}. 
To show that the neural network is both interpolating and extrapolating well we show its 
behaviour on random trajectories in phase-space in Section~\ref{sec:trajectories}. 
Our conclusions are presented in Section~\ref{sec:conclusion} 
and some discussion of specific details are collected in the appendices so as not to distract from the main discussion.

We provide code to reproduce the methodology detailed in this article \cite{fame_repo}.

@@

\subsection{Infrared divergences and dipole factorisation formula}
It is well known that in soft and collinear limits the matrix element in $n+1$-body phase-space factorises into a singular factor and a reduced matrix element in $n$-body phase-space \cite{Altarelli:1977zs,BASSETTO1983201}.
This factorisation was used by Catani and Seymour \cite{Catani:1996vz} to construct subtraction terms for the real radiation part of a NLO calculation.
They introduced a factorisation formula with universal dipoles that smoothly interpolates 
between the soft and collinear limits to capture the singular structure in these regions of phase-space.
The dipole factorisation formula can be written schematically as
\begin{eqnarray}\label{eq:factorisation}
    |\mathcal{M}_{n+1}|^{2} \rightarrow |\mathcal{M}_{n}|^{2} \otimes \mathbf{V}_{ij,k} \, ,
\end{eqnarray}
where $\mathbf{V}_{ij,k}$ is a process independent, singular factor. It depends on the momenta and quantum numbers (colour and spin) of partons $i,j,k$, where $i$ is the emitter parton, 
$j$ is the emitted parton, and $k$ is the spectator parton.
For singly unresolved limits, this factorisation isolates all the divergent behaviour in $\mathbf{V}_{ij,k}$ and the factor $|\mathcal{M}_{n}|^{2}$ is free of divergences, which makes 
it more amenable to emulation through a neural network.
The dipole factorisation formula forms the basis of our fitting ansatz which we present in detail in Section~\ref{sec:ansatz}.

@@

\paragraph*{Training, validation, and testing datasets}
In Ref. \cite{Badger:2020uow}, the authors used 500k training samples to train their models.
For a fair comparison of our respective methods, we follow their methodology closely by constructing our models to be as close as possible to theirs and generate training and testing data by using code from their project repository \cite{n3jet_repo}\footnote{for our main results we generate all data using our own code}.
Our model architectures will not be identical due to the difference in our methods, details of our model architecture are given in Section~\ref{sec:architecture}.
Results of this comparison are presented in Section~\ref{sec:n3jet_comparison}.

@@

Each model in the ensemble is initialised with different weights according to the `Glorot uniform' distribution, and trained on the same but randomly shuffled dataset, resulting in models that have been exposed to different distributions of the training data.
After sufficient training, each model will have a distinct set of parameters that have similar predictive power.
The prediction for the matrix element is then the mean of the outputs of all models in the ensemble\footnote{if not explicitly stated, all references to neural network predictions from henceforth refers to the prediction from the ensemble of networks}.
Taking the mean will give a more accurate and robust prediction as averaging over the different models will reduce variance due to over/underfitting in the training phase.
We choose to have 20 models in our ensembles because we begin to see diminishing returns in the accuracy of per-point predictions after this.
Another advantage of training an ensemble of models is that we have a measure of uncertainty, due to the neural networks, on the model predictions by calculating the standard error of the mean.
That is we take the standard deviation of predictions across the ensemble and divide by the square root of number of models in the ensemble. This would not be possible with just a single model.

@@

By increasing the size of the training datasets we aim to expose the neural network to more samples of the phase-space, thereby increasing accuracy on predictions on as much of the phase-space as possible.
Along with increasing the number of training samples, the neural network architecture has been expanded to include more hidden nodes and hidden layers.
The extra hidden nodes and layers introduces more parameters into the model allowing the neural network to utilise the additional data. 
We found that to get good performance, we had to balance the size of the training dataset used and the size of the network. i.e. a small network is not expected to capture all the variations in a large dataset as easily as a larger network would, due to the smaller number of parameters available to the network.
Of course, we also had to consider more physical constraints such as the time spent on training the neural networks which limits both the size of training datasets and architecture.
In Figure~\ref{fig:error_500k_vs_40m}, we show the improvements in accuracy of our main NNs compared to the smaller NNs from Section~\ref{sec:n3jet_comparison}.
Although the training and testing data for the smaller NNs are not identical due to differences in code used to generate the sets\footnote{although RAMBO is used for phase-space generation in both works, the selection criteria is different (JADE algorithm vs $k_{t}$ algorithm)}, for this comparison it suffices to show that the larger NNs are orders of magnitude more accurate than the smaller NNs.

@@

\subsection{Random trajectories}\label{sec:trajectories}
As another test of our emulator we assess the accuracy of our predictions on random phase-space trajectories.
These random phase-space trajectories are generated by connecting two random points in phase-space continuously without excluding any region of phase-space.
This presents an interesting and challenging test of the interpolation and extrapolation abilities of the NNs as some parts of the trajectories may lie outside of the phase-space region of the training datasets.
We show the results for 5-jet trajectories as this is the highest multiplicity we considered, predictions on lower multiplicities are better-behaved.
We investigated 50 different random trajectories\footnote{the corresponding plots are available in a document attached to this submission}. For the discussion in this article 
we chose one that contains many interesting features, namely the matrix elements span many orders of magnitudes and there are distinct peaks in the trajectory.
In Figure~\ref{fig:random_trajectory} we show the predictions by the three 5-jet models trained on data with different global phase-space cuts for this trajectory.
Left column shows the actual matrix element prediction and right column shows the ratios of the prediction to {\NJet}.
We analysed the predictions for the 50 random trajectories and measured the fraction of their length where the accuracy falls within given intervals.
Table~\ref{table:trajectory_predictions} shows the result for the regions of phase-space where training data was available, and those falling beyond the data available to the model.

@@

\section{Conclusion}\label{sec:conclusion}
In this article we presented a new strategy to emulate matrix elements using a neural network. By leveraging the knowledge of 
the factorisation properties of the matrix elements our model is able to extrapolate well outside of its training range. 
We showed that using this method we obtain significantly improved per-point accuracy than obtained in previous works. 
We also showed that the per-point accuracy of the model is not significantly affected by the generation cut for the training, which means that
it would be possible to train our emulator on very inclusive cuts, allowing them to be applied in a multitude of settings.

@@

Our method was demonstrated in this article using a tree-level process, but it could be generalised to higher order matrix elements by adapting the set 
of ingredients made available to the network for the interpolation. Specifically, we can include extra terms into our ansatz to help capture additional divergences at higher orders. We look forward to applying this method to such cases.

>>>>

\section{Introduction}
\label{sec:intro}
The capacity of neural networks to approximate intricate functions have already been 
used to provide fast calculations of production cross-sections~\cite{Otten:2018kum,Buckley:2020bxg}.
In this chapter we investigate whether NNs can approximate production cross-sections more differentially by replacing computationally expensive 
matrix element calculations. 
The challenge with this endeavour is that matrix elements are plagued with numerous divergences that arise from infrared divergences.
In previous works \cite{Badger:2020uow,Aylett-Bullock:2021hmo} a combination of individual neural networks were used 
to approximate matrix elements. In order to deal with the complex structure of the matrix elements 
the authors of these studies divided the phase-space into sectors according to the infrared singularities and trained networks on these sectors,
thereby limiting the complexity of the fit by isolating a single divergence per sector. All sectors were then combined to make the final prediction.
While the authors of the initial study for electron-positron annihilation showed good agreement between the total cross-section and histogrammed distributions both at LO and NLO,
we note that the accuracy of the interpolation at the level of individual points was a lot worse than when averaged in the 
histograms. In addition, the performance of the extrapolation outside of its training phase-space 
(i.e. more singular configurations than those considered to fit the model) is problematic.

In this chapter we present a different approach to the emulation of matrix elements that incorporates the factorisation 
properties of the matrix elements in the interpolation model and therefore is able to safely 
extrapolate the matrix element in regions more singular than that covered by the data used for the training of the emulator. We find
that our interpolator also displays a much improved pointwise accuracy.

This chapter is organised as follows. 
Section~\ref{sec:fitting} introduces our factorisation-aware 
deep neural network model, Section~\ref{sec:results} showcases our results with a comparison with the 
model from Ref.~\cite{Badger:2020uow} in Section~\ref{sec:n3jet_comparison}, and an analysis of the performance of our model in Section~\ref{sec:mainresult}. 
To show that the neural network is both interpolating and extrapolating well we show its 
behaviour on random trajectories in phase-space in Section~\ref{sec:trajectories}. 
Our conclusions are presented in Section~\ref{sec:conclusion} 
and some discussion of specific details are collected in the appendices so as not to distract from the main discussion.

Computer code to reproduce the methodology detailed in this chapter is provided at \cite{fame_repo}.

@@

\subsection{Infrared divergences and dipole factorisation formula}
It is well known that in soft and collinear limits the matrix element in $(n+1)$-body phase-space factorises into a singular factor and a reduced matrix element in $n$-body phase-space \cite{Altarelli:1977zs,BASSETTO1983201} (Section~\ref{sec:me_factorisation}).
This factorisation was used by Catani and Seymour \cite{Catani:1996vz} to construct subtraction terms for the real radiation part of a NLO calculation (Section~\ref{sec:CS_dipoles}).
They introduced a factorisation formula with universal dipoles that smoothly interpolates 
between the soft and collinear limits to capture the singular structure in these regions of phase-space.
The dipole factorisation formula can be written schematically as
\begin{eqnarray}\label{eq:factorisation}
    |\mathcal{M}_{n+1}|^{2} \rightarrow |\mathcal{M}_{n}|^{2} \otimes \mathbf{V}_{ij,k} \, ,
\end{eqnarray}
where $\mathbf{V}_{ij,k}$ is a process independent, singular factor. It depends on the momenta and quantum numbers (colour and spin) of partons $i,j,k$, where $i$ is the emitter parton, 
$j$ is the emitted parton, and $k$ is the spectator parton.
For singly unresolved limits, this factorisation isolates all the divergent behaviour in $\mathbf{V}_{ij,k}$ and the factor $|\mathcal{M}_{n}|^{2}$ is free of divergences, which makes 
it more amenable to emulation through a neural network.
The dipole factorisation formula forms the basis of our fitting ansatz which we present in detail in Section~\ref{sec:ansatz}.

@@

\paragraph*{Training, validation, and testing datasets}
In Ref. \cite{Badger:2020uow}, the authors used 500k training samples to train their models.
For a fair comparison of our respective methods, we follow their methodology closely by constructing our models to be as close as possible to theirs and generate training and testing data by using code from their project repository \cite{n3jet_repo}\footnote{For our main results we generate all data using our own code}.
Our model architectures will not be identical due to the difference in our methods, details of our model architecture are given in Section~\ref{sec:architecture}.
Results of this comparison are presented in Section~\ref{sec:n3jet_comparison}.

@@

Each model in the ensemble is initialised with different weights according to the `Glorot uniform' distribution, and trained on the same but randomly shuffled dataset, resulting in models that have been exposed to different distributions of the training data.
After sufficient training, each model will have a distinct set of parameters that have similar predictive power.
The prediction for the matrix element is then the mean of the outputs of all models in the ensemble\footnote{If not explicitly stated, all references to neural network predictions henceforth refers to the prediction from the ensemble of networks}.
Taking the mean will give a more accurate and robust prediction as averaging over the different models will reduce variance due to over/underfitting in the training phase.
We choose to have 20 models in our ensembles because we begin to see diminishing returns in the accuracy of per-point predictions after this.
Another advantage of training an ensemble of models is that we have a measure of uncertainty, due to the neural networks, on the model predictions by calculating the standard error of the mean.
That is we take the standard deviation of predictions across the ensemble and divide by the square root of number of models in the ensemble. This would not be possible with just a single model.

@@

By increasing the size of the training datasets we aim to expose the neural network to more samples of the phase-space, thereby increasing accuracy on predictions on as much of the phase-space as possible.
Along with increasing the number of training samples, the neural network architecture has been expanded to include more hidden nodes and hidden layers.
The extra hidden nodes and layers introduces more parameters into the model allowing the neural network to utilise the additional data. 
We found that to get good performance, we had to balance the size of the training dataset used and the size of the network. i.e. a small network is not expected to capture all the variations in a large dataset as easily as a larger network would, due to the smaller number of parameters available to the network.
Of course, we also had to consider more physical constraints such as the time spent on training the neural networks which limits both the size of training datasets and architecture.
In Figure~\ref{fig:error_500k_vs_40m}, we show the improvements in accuracy of our main NNs compared to the smaller NNs from Section~\ref{sec:n3jet_comparison}.
Although the training and testing data for the smaller NNs are not identical due to differences in code used to generate the sets\footnote{Although RAMBO is used for phase-space generation in both works, the selection criteria is different (JADE algorithm vs $k_{t}$ algorithm).}, for this comparison it suffices to show that the larger NNs are orders of magnitude more accurate than the smaller NNs.

@@

\subsection{Random trajectories}\label{sec:trajectories}
As another test of our emulator we assess the accuracy of our predictions on random phase-space trajectories.
These random phase-space trajectories are generated by connecting two random points in phase-space continuously without excluding any region of phase-space.
This presents an interesting and challenging test of the interpolation and extrapolation abilities of the NNs as some parts of the trajectories may lie outside of the phase-space region of the training datasets.
We show the results for 5-jet trajectories as this is the highest multiplicity we considered, predictions on lower multiplicities are better-behaved.
We investigated 50 different random trajectories. For the discussion in this section
we chose one that contains many interesting features, namely the matrix elements span many orders of magnitudes and there are distinct peaks in the trajectory.
In Figure~\ref{fig:random_trajectory} we show the predictions by the three 5-jet models trained on data with different global phase-space cuts for this trajectory.
Left column shows the actual matrix element prediction and right column shows the ratios of the prediction to {\NJet}.
We analysed the predictions for the 50 random trajectories and measured the fraction of their length where the accuracy falls within given intervals.
Table~\ref{table:trajectory_predictions} shows the result for the regions of phase-space where training data was available, and those falling beyond the data available to the model.

@@

\section{Conclusion}\label{sec:conclusion}
In this chapter we presented a new strategy to emulate matrix elements using a neural network. By leveraging the knowledge of 
the factorisation properties of the matrix elements our model is able to extrapolate well outside of its training range. 
We showed that using this method we obtain significantly improved per-point accuracy than obtained in previous works. 
We also showed that the per-point accuracy of the model is not significantly affected by the generation cut for the training, which means that
it would be possible to train our emulator on very inclusive cuts, allowing them to be applied in a multitude of settings.

@@

Our method was demonstrated in this chapter using a tree-level process, but it could be generalised to higher order matrix elements by adapting the set 
of ingredients made available to the network for the interpolation. This is explicitly shown in the next chapter for the same process at NLO QCD.
